Bayesian Nonparametric Motor-skill Representations \\ for Efficient Learning of Robotic Clothing Assistance

### 2016-11-27 Paper/paper.tex
[major] complete 3rd draft of paper
[minor] edit minor comments for camera ready paper

\begin{abstract}
Motor-skill learning for complex robotic tasks is a challenging problem due to the high task variability. Robotic clothing assistance is one such challenging problem that can greatly improve the quality-of-life for the elderly and disabled. In this study, we propose a data-efficient representation to encode task specific motor-skills of the robot using Bayesian nonparametric latent manifold learning. We implement our framework in a practical setting with a dual-arm robot performing clothing tasks. We demonstrate that performing motor-skills learning in such task specific latent spaces outperforms learning in the high-dimensional joint configuration space of the robot. Furthermore, our framework can also be used as a tool for learning from demonstration to impart novel skills to the robot.
\end{abstract}

@@

Enabling robots to interact with the environment and learning motor skills to perform complex tasks is one of the major goals of the AI community. However, existing methods require large number of interactions to learn optimal behavior which is not suitable in practical scenarios such as in the field of Assistive Robotics. Clothing assistance is one such task which is a necessity in the daily life of the elderly and disabled people. Design of a robotic framework involves a motor skills learning framework that can detect and adapt to various failure scenarios. Tamei \emph{et al.}~\cite{tamei} proposed a reinforcement learning (RL) framework for clothing assistance with a dual-arm robot to cloth a soft mannequin with a T-shirt. For tractable learning time, policy update was done using finite difference policy gradient applied to a small subset of policy parameters. This constrained generalization capability to very environmental settings such as major changes in the subject's posture or using different clothing article.

To address this problem, we need to consider an alternate representation that is more flexible and suitable for robust learning of motor-skills. An efficient approach is the use of dimensionality reduction (DR) along with reinforcement learning (RL). There have been several studies in this direction in the recent years. Bitzer \emph{et al.}~\cite{bitzer} proposed a learning framework where DR was used as a preprocessing step to extract the problem structure. However, this study relies on a maximum-a-posteriori (MAP) estimate of the latent space and can overfit to the training data. Luck \emph{et al.}~\cite{luck} proposed a policy search framework that inherently combines RL and DR wherein the latent space is learned based on the reward signals. However, this framework relies on the use of linear dimensionality reduction which limits applicability to complex tasks.

In this study, we propose an efficient representation of motor skills that relies on the use of Bayesian Gaussian Process Latent Variable Model (BGPLVM)~\cite{bgplvm}. We hypothesize that the latent space learned using BGPLVM is an efficient search space to learn novel motor-skills for complex robotics tasks such as clothing assistance. Furthermore, the mapping from latent space to data space is modeled using a Gaussian Process (GP)~\cite{gp} allowing data-efficient learning. This is crucial to ensure the robot can generalize well to unseen environmental settings within few interactions making it suitable for a feasible social implementation of robotic clothing assistance. We apply our proposed method to a clothing assistance setup as shown in Figure~\ref{figure:overview}. We demonstrate that performing policy search in this latent space outperforms learning in the high-dimensional action space. We further present the design of a real-time controller from the BGPLVM latent space that can be used as a tool for Learning from Demonstration (LfD). The experimental results indicate a promising reinforcement learning framework that can be used for challenging tasks such as robotic clothing assistance.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.2\textheight]{overview.png}
  \caption{Overview of proposed framework. Figure on the left shows the setup for clothing assistance and the figure on the right shows the latent space encoding the motor skills. This latent space can be used a tool for Learning from Demonstration (LfD) and also for data-efficient reinforcement learning.}
  \label{figure:overview}
\end{figure}
Motor skills for clothing assistance are given by high dimensional joint angle trajectories of the dual-arm robot. The robot also has to maintain several task space constraints such as coupling with a clothing article along with safe human-robot interaction as shown in Figure~\ref{figure:overview}. To address these problems, we propose the use of BGPLVM for learning a low-dimensional latent space through a nonlinear mapping to the kinematic space. There several motivating factors for this choice. The Bayesian treatment avoids over fitting to the training data and the use of an ARD kernel in the GP mapping leads to inference of the inherent dimensionality to encode motor-skills.

@@

Bayesian Gaussian Process Latent Variable Model (BGPLVM) is a dimensionality reduction technique proposed by Titsias \emph{et al.}~\cite{bgplvm}  derived from the probabilistic model where the observations, $\mathbf{Y} = \{\mathbf{y}_{i} \in \mathbb{R}^D\}_{n=1}^N$, are assumed to be generated from latent inputs $\mathbf{X} = \{\mathbf{x}_{i} \in \mathbb{R}^q\}_{n=1}^N$ through a noisy process,
\begin{equation}
  \mathbf{y}_i = f(\mathbf{x}_i) + \epsilon,~\epsilon \sim \mathcal{N}(\mathbf{0},\beta^{-1}\mathbf{I})
\end{equation}
The mapping function $f$ is modeled using a Gaussian Process (GP) suitable for performing non-linear dimensionality reduction with use of a non-linear kernel function. The marginal likelihood for the generative model is given by:
\begin{equation}
  \label{eqn:marginallikelihood}
  p(\mathbf{Y}|\mathbf{X},\Phi) = \prod_{d=1}^{D} \mathcal{N}(\mathbf{y}_{:,d}|\mathbf{0},\mathbf{K}+\beta^{-1}I)\\
\end{equation}
where $\mathbf{X},\Phi$ are the unknown latent points and hyper parameters for the GP mapping that need to be inferred. $\mathbf{K}$ is the kernel matrix constructed from the latent points. In the generative model, the latent positions need to be marginalized out for having a purely Bayesian treatment. Titsias \emph{et al.}~\cite{bgplvm} proposed a variational inference approach to compute a tractable lower bound for the marginalization thereby inferring a posterior distribution on the latent positions rather than a MAP estimate.

@@

We consider Dynamic Movement Primitives (DMP)~\cite{dmp} as the policy representation. It is combination of a point attractor dynamical system and a non-linear forcing term learned using Locally Weighted Regression (LWR)~\cite{lwr}.
%\begin{equation}
%  \begin{array}{c}
%    \tau \ddot{x} = K(g-x) -D\dot{x} + (g - x_0)f\\~\\
%    \displaystyle f(s) = \frac{\sum_i w_i \psi_i(s)s}{\sum_i \psi_i(s)}, \text{ where } \tau \dot{s} = - \alpha s
%  \end{array}
%\end{equation}
We train a DMP on one of the latent trajectories and the LWR wight coefficients are used as the policy parameters. The cost function for policy improvement is designed in the high-dimensional action space. In the current setting, we obtain a demonstration for the unseen posture and consider this as the optimized trajectory to be achieved. The optimized trajectory is efficiently encoded by via-points extracted using the minimum jerk criterion~\cite{via} that the robot needs to pass through. The cost function is given by the sum of all errors between the reconstructed robot trajectory and the desired via-points:
\begin{equation}
  R(\tau) = \sum_{i = 1}^{n_{\text{dims}}} \sum_{j = 1}^{n_{\text{via}}} (V_{i,j} - x^{recons}_i(t_{i,j}))
\end{equation}
where $R(\tau)$ is the total reward for trajectory $\tau$, $V_{i,j}$ is the $j^{\text{th}}$ via point of $i^{\text{th}}$ dimension and $x^{recons}_i(t)$ is the value at time $t$ for $i^{\text{th}}$ dimension of reconstructed trajectory.
\section{Results}
\label{section:results}

@@

In this section, we present the applicability of BGPLVM for encoding clothing skills. The evaluation dataset contains 4 clothing trials that were obtained for 4 different postures of the mannequin wherein the elevation of the arms were varied forming different angles $\{65^o, 70^o, 75^o, 80^o\}$ with the mannequin's body. The clothing trial for $75^o$ was left out as the test data and latent variable models were trained for the remaining 3 clothing trials. The ARD weights on training resulted in 3 active latent dimensions for the joint angle scenario and 2 latent dimensions for the end effector scenario. This implies that the motor skills encoded varies depending the input action space. The latent space learned for the joint angle scenario is shown in Figure~\ref{figure:latentspace}.

@@

We further demonstrate the effectivity of performing policy search in the BGPLVM latent space by comparison with search in the high-dimensional action space and in latent spaces learned using PCA and GPLVM. For the comparison, the policy search method, policy representation and reward function are the same and the main difference is the search space. CEM was run for 50 iterations where 50 rollouts were performed in each iteration and the top 5 rollouts were retained for computing the mean and variance of the policy parameters. Figure~\ref{figure:rlcurve} shows the learning curves for all formulations and the BGPLVM scenario performs better than all other search spaces. This validates our hypothesis that encoding motor-skills using Bayesian nonparametric latent variable models is an effective parametrization for learning.

>>>>

\begin{abstract}
Motor-skill learning for complex robotic tasks is a challenging problem due to the high task variability. Robotic clothing assistance is one such challenging problem that can greatly improve the quality-of-life for the elderly and disabled. In this study, we propose a data-efficient representation to encode task specific motor-skills of the robot using Bayesian nonparametric latent space learning. We implement our framework in a practical setting with a dual-arm robot performing clothing assistance tasks. We demonstrate that performing motor-skills learning in such task specific latent spaces outperforms learning in the high-dimensional joint configuration space of the robot. Furthermore, our framework can also be used as a tool for learning from demonstration to impart novel skills to the robot.
\end{abstract}

@@

Enabling robots to interact with the environment and learning motor skills to perform complex tasks is one of the major goals of the AI community. However, existing methods require large number of interactions to learn optimal behavior which is not suitable in practical scenarios such as in the field of Assistive Robotics. Clothing assistance is one such task which is a necessity in the daily life of the elderly and disabled people. Design of a robotic framework involves a motor skills learning framework that can detect and adapt to various failure scenarios. Tamei \emph{et al.}~\cite{tamei} proposed a reinforcement learning (RL) framework for clothing assistance with a dual-arm robot to cloth a soft mannequin with a T-shirt. For tractable learning time, policy update was done using finite difference policy gradient applied to a small subset of policy parameters. This constrained the generalization capability of the algorithm to very different environmental settings such as major changes in the subject's posture or using different clothing articles.

To address this problem, we need to consider an alternate representation that is more flexible and suitable for robust learning of motor-skills. An efficient approach is the use of dimensionality reduction (DR) along with reinforcement learning (RL). There have been several studies in this direction in recent years. Bitzer \emph{et al.}~\cite{bitzer} proposed a learning framework where DR was used as a preprocessing step to extract problem structure. However, this study relies on a maximum-a-posteriori (MAP) estimate of the latent space and can overfit to the training data. Luck \emph{et al.}~\cite{luck} proposed a policy search framework that inherently combines RL and DR wherein the latent space is learned based on the reward signals. However, this framework relies on the use of linear dimensionality reduction which limits applicability to complex tasks.

In this study, we propose an efficient representation of motor skills that relies on the use of Bayesian Gaussian Process Latent Variable Model (BGPLVM)~\cite{bgplvm}. We hypothesize that the latent space learned using BGPLVM is an efficient search space to learn novel motor-skills for complex robotic tasks such as clothing assistance. The mapping from latent space to data space is modeled using a Gaussian Process (GP)~\cite{gp} allowing data-efficient learning. This is crucial to ensure the robot can generalize well to unseen environmental settings within few interactions making it suitable for a feasible social implementation of robotic clothing assistance. We apply our proposed method to a clothing assistance setup as shown in Figure~\ref{figure:overview}. We demonstrate that performing policy search in this latent space outperforms learning in the high-dimensional action space. We further present the design of a real-time controller from the BGPLVM latent space that can be used as a tool for Learning from Demonstration (LfD). The experimental results indicate a promising reinforcement learning framework that can be used for challenging tasks such as robotic clothing assistance.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth, height=0.2\textheight]{overview.png}
  \caption{Overview of proposed framework. Figure on the left shows the setup for clothing assistance and the figure on the right shows the latent space encoding the motor skills. This latent space can be used as a tool for Learning from Demonstration (LfD) and also for data-efficient reinforcement learning.}
  \label{figure:overview}
\end{figure}
Motor skills for clothing assistance are given by high dimensional joint angle trajectories of the dual-arm robot. The robot also has to maintain several task space constraints such as coupling with a clothing article along with safe human-robot interaction as shown in Figure~\ref{figure:overview}. To address these problems, we propose the use of BGPLVM for learning a low-dimensional latent space through a nonlinear mapping to the kinematic space. There are several motivating factors for this choice. The Bayesian treatment avoids overfitting to the training data and the use of an ARD kernel in the GP mapping leads to inference of the inherent dimensionality to encode motor-skills.

@@

Bayesian Gaussian Process Latent Variable Model (BGPLVM) is a dimensionality reduction technique proposed by Titsias \emph{et al.}~\cite{bgplvm}  derived from the probabilistic model where the observations, $\mathbf{Y} = \{\mathbf{y}_{i} \in \mathbb{R}^D\}_{n=1}^N$, are assumed to be generated from latent inputs $\mathbf{X} = \{\mathbf{x}_{i} \in \mathbb{R}^q\}_{n=1}^N$ through a noisy process,
\begin{equation}
  \mathbf{y}_i = f(\mathbf{x}_i) + \epsilon,~\epsilon \sim \mathcal{N}(\mathbf{0},\beta^{-1}\mathbf{I})
\end{equation}
The mapping function $f$ is modeled using a Gaussian Process (GP) suitable for performing non-linear dimensionality reduction with use of a non-linear kernel function. The marginal likelihood for the generative model is given by:
\begin{equation}
  \label{eqn:marginallikelihood}
  p(\mathbf{Y}|\mathbf{X},\Phi) = \prod_{d=1}^{D} \mathcal{N}(\mathbf{y}_{:,d}|\mathbf{0},\mathbf{K}+\beta^{-1}I)\\
\end{equation}
where $\mathbf{X},\Phi$ are the unknown latent points and hyperparameters for the GP mapping that need to be inferred. $\mathbf{K}$ is the kernel matrix constructed from the latent points. In the generative model, the latent positions need to be marginalized out for having a purely Bayesian treatment. Titsias \emph{et al.}~\cite{bgplvm} proposed a variational inference approach to compute a tractable lower bound for the marginalization thereby inferring a posterior distribution on the latent positions rather than a MAP estimate.

@@

We consider Dynamic Movement Primitives (DMP)~\cite{dmp} as policy representation. It is the combination of a point attractor dynamical system and a non-linear forcing term learned using Locally Weighted Regression (LWR)~\cite{lwr}.
%\begin{equation}
%  \begin{array}{c}
%    \tau \ddot{x} = K(g-x) -D\dot{x} + (g - x_0)f\\~\\
%    \displaystyle f(s) = \frac{\sum_i w_i \psi_i(s)s}{\sum_i \psi_i(s)}, \text{ where } \tau \dot{s} = - \alpha s
%  \end{array}
%\end{equation}
We train a DMP on one of the latent trajectories and the LWR weight coefficients are used as policy parameters. The cost function for policy improvement is designed in the high-dimensional action space. In the current setting, we obtain a demonstration for the unseen posture and consider this as the optimized trajectory to be achieved. The optimized trajectory is efficiently encoded by via-points extracted using the minimum jerk criterion~\cite{via} that the robot needs to pass through. The cost function is given by the sum of all errors between the reconstructed robot trajectory and the desired via-points:
\begin{equation}
  R(\tau) = \sum_{i = 1}^{n_{\text{dims}}} \sum_{j = 1}^{n_{\text{via}}} (V_{i,j} - x^{recons}_i(t_{i,j}))
\end{equation}
where $R(\tau)$ is the total reward for trajectory $\tau$, $V_{i,j}$ is the $j^{\text{th}}$ via point of $i^{\text{th}}$ dimension and $x^{recons}_i(t)$ is the value at time $t$ for $i^{\text{th}}$ dimension of reconstructed trajectory.
\section{Results}
\label{section:results}

@@

In this section, we present the applicability of BGPLVM for encoding clothing skills. The evaluation dataset contains 4 clothing trials that were obtained for 4 different postures of the mannequin wherein the elevation of the arms were varied forming different angles $\{65^o, 70^o, 75^o, 80^o\}$ with the mannequin's body. The clothing trial for $75^o$ was left out as the test data and latent variable models were trained for the remaining 3 clothing trials. The ARD weights on training resulted in 3 active latent dimensions for the joint angle scenario and 2 latent dimensions for the end effector scenario. This implies that the encoded motor skills varies depending on the input action space. The latent space learned for the joint angle scenario is shown in Figure~\ref{figure:latentspace}.

@@

We further demonstrate the effectivity of performing policy search in the BGPLVM latent space by comparison with search in the high-dimensional action space and in latent spaces learned using PCA and GPLVM. CEM was run for 50 iterations where 50 rollouts were performed in each iteration. The rollouts were generated from DMP with random sampling of LWR weight parameters. The top 5 rollouts were retained for updating the mean and variance of policy parameters at each iteration. Figure~\ref{figure:rlcurve} shows the learning curves for all formulations and the BGPLVM scenario performs better than all other search spaces. PCA suffers from linearity assumptions and as GPLVM relies on a MAP estimate of the latent space it could be overfitting to the training data. This validates our hypothesis that encoding motor-skills using Bayesian nonparametric latent variable models is an effective parametrization for learning.
